A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-04-18 inst/manuscript/manuscript.Rmd
minor review updates

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between these design-based inference and model-based inference is small GRTS samples are used but large when SRS samples are used. This suggests that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

Spatial data are data that have some sort of spatial index --usually this index is specified via coordinates. @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling
design assigns a probability of selection to each sample (subset of population
units). Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. The inclusion probability of a population unit follows by summing each sample's probability of selection over all samples that contain the population unit. Inclusion probabilities are later used to estimate population parameters.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata @walvoort2010r, the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or on private land that the researcher does not have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. First the bounding box of the domain is split up into four distinct, equally sized squares called level-one cells. Each level-one cell is randomly assigned a level-one address of 0, 1, 2, or 3. The set of level-one cells is denoted by $\mathcal{A}_1$ and defined as $\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}$. Within each level-one cell, the inclusion probability for each population unit is summed, and if any of these sums exceed one, a second level of cells is added. Then each level-one cell is split into four distinct, equally sized squares called level-two cells. Each level-two cell is randomly assigned a level-two address of 0, 1, 2, or 3. The set of level-two cells is denoted by $\mathcal{A}_2$ and defined as $\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}$. The inclusion probabilities within each level-two cell are summed, and if any of these sums exceed one, a third level of cells is added. This process continues for $k$ steps, until all level-$k$ cells have inclusion probability sums no larger than one. Then $\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}$.

After determining $\mathcal{A}_k$, it is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set \linebreak $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space. Next each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each of these sample points falls on some population unit's line segment, and thus that population unit is selected in the sample. For further details regarding the GRTS algorithm, see @stevens2004spatially.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimator for $\tau$, denoted by $\hat{\tau}_{ht}$, is is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align}
where $z_i$ is the value of the $i$th population unit in the sample, $\pi_i$ is the inclusion probability of the $i$th population unit in the sample, and $n$ is the sample size. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

@@

Finite population block kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FPBK next -- for more details, see @verhoef2008spatial. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\ref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), isotropic (independent of direction), and decay with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$. @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is often called the partial sill, $\sigma^2_{2}$ is often called the nugget, and $\phi$ is often called the range.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\delta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\delta}}_{reml}$, the REML estimates of $\bm{\delta}$. Then $\bm{\hat{\beta}}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\delta}}_{reml}$.

@@

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. We used real data from the 2012 National Lakes Assessment to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the real data, we were in control of only the random sampling process.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the the halved average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

@@

Mean bias was nearly zero for all four sampling-inference combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for mean bias in all 36 simulation scenarios are provided in the supporting information.

We define the relative RMSE as a ratio with numerator given by the RMSE for a sampling-inference combination and the denominator given by the RMSE for SRS-DB. Relative RMSEs for the random location layout are provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%"), the four sampling-inference combinations have approximately equal RMSE. In these scenarios, using GRTS sampling or model-based inference does not generally increase efficiency compared to SRS-DB. When there is spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 50%" and "DRE%: 90%"), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB, SRS-MB, and finally SRS-DB, though the difference in relative RMSE among GRTS-MB, GRTS-DB, and SRS-MB is small.  As the strength of spatial covariance increases, the gap in RMSE between SRS-DB and the other sampling-inference combinations widens. Finally we note that when there is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB, suggesting that the lack of efficiency from SRS is largely mitigated by model-based inference. These RMSE conclusions are similar to those observed in the grid location layout, so we omit a figure and discussion regarding the grid location layout here. Tables for RMSE in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}

>>>>

---
title: A comparison of design-based and model-based approaches for finite population spatial sampling and inference.
author:
  - name: Michael Dumelle
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    affiliation: STLAW
  - name: Jay M. Ver Hoef
    affiliation: NOAA
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Lisa Madsen
    affiliation: OSU
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Mathematics, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
footnote:
    code: 1
    text: "Corresponding Author: Michael Dumelle (Dumelle.Michael@epa.gov)"
abstract: |
  1.   The design-based and model-based approaches to frequentist statistical inference rest on fundamentally different foundations. In the design-based approach, inference relies on random sampling. In the model-based approach, inference relies on distributional assumptions. We compare the approaches for finite population spatial data.
  2. We provide relevant background for the design-based and model-based approaches and then study their performance using simulated and real data from the United States Environmental Protection Agency's 2012 National Lakes Assessment. A variety of sample sizes, location layouts, dependence structures, and response types are considered. The population mean is the parameter of interest and performance is measured using statistics like bias, squared error, and interval coverage.
  3. When studying the simulated and real data, we found that regardless of the strength of spatial dependence in the data, the generalized random tessellation stratified (GRTS) algorithm, which explicitly incorporates spatial locations into sampling, tends to outperform the simple random sampling (SRS) algorithm, which does not explicitly incorporate spatial locations into sampling. We also found that model-based inference tends to outperform design-based inference, even for skewed data where the model-based distributional assumptions are violated. The performance gap between design-based inference and model-based inference is small when GRTS samples are used but large when SRS samples are used, suggesting that the sampling choice (whether to use GRTS or SRS) is most important when performing design-based inference.
  4. There are many benefits and drawbacks to the design-based and model-based approaches for finite population spatial sampling and inference that practitioners must consider when choosing between them. We provide relevant background contextualizing each approach and study their properties in a variety of scenarios, making recommendations for use based on the practitioner's goals.
journal: "Methods in Ecology and Evolution"
date: "`r Sys.Date()`"
bibliography: mybibfile.bib
linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
preamble: >
  \usepackage{bm}
  \usepackage{bbm}
  \usepackage{color}
  \DeclareMathOperator{\var}{{var}}
  \DeclareMathOperator{\cov}{{cov}}
  \usepackage{caption}
  \usepackage{subcaption}
  \usepackage{setspace}
  \usepackage{fancyhdr}
  \pagestyle{fancy}
  \fancyhf{}
  \chead{Spatial design-based vs model-based}
  \doublespacing
output: rticles::elsevier_article
---

@@

Spatial data are data that have some sort of spatial index (usually specified via coordinates). @de1990model and @brus1993design give early comparisons of design-based and model-based approaches for spatial data, quashing the belief that design-based approaches could not be used for spatially correlated data. Since then, there have been several general comparisons between design-based and model-based approaches for spatial data [@brus1997random; @verhoef2002sampling; @verhoef2008spatial; @brus2021statistical]. @cooper2006sampling reviews the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. In addition to @cooper2006sampling, there has been substantial research and development into estimators that use both design-based and model-based principles (see e.g., @sterba2009alternative and @cicchitelli2012model, and for Bayesian approaches, see @chan2020bayesian and @hofman2021many).

@@

The design-based approach assumes the population is fixed. Randomness is incorporated via the selection of population units according to a sampling design. A sampling
design assigns a probability of selection to each sample (subset of population
units). Some examples of commonly used sampling designs include simple random sampling, stratified random sampling, and cluster sampling. The inclusion probability of a population unit is calculated by summing each sample's probability of selection over all samples that contain the population unit. Inclusion probabilities are often when selecting samples and estimating population parameters.

@@

We previously mentioned that the design-based approach can be used to select spatially balanced samples. Spatially balanced samples are useful because parameter estimates from these samples tend to vary less than parameter estimates from samples lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @wang2013design; @benedetti2017spatiallyreview]. The first spatially balanced sampling algorithm to see widespread use was the generalized random tessellation stratified (GRTS) algorithm [@stevens2004spatially]. After the GRTS algorithm was developed, several other spatially balanced sampling algorithms emerged, including stratified sampling with compact geographical strata [@walvoort2010r], the local pivotal method [@grafstrom2012spatially; @grafstrom2018spatially], spatially correlated Poisson sampling [@grafstrom2012spatiallypoisson], balanced acceptance sampling [@robertson2013bas], within-sample-distance sampling [@benedetti2017spatially], and Halton iterative partitioning sampling [@robertson2018halton]. In this manuscript, we select spatially balanced samples using the GRTS algorithm because it is readily available in the `spsurvey` \textbf{\textsf{R}} package [@dumelle2022spsurvey] and naturally accommodates finite and infinite sampling frames, unequal inclusion probabilities, and replacement units. Replacement units are additional population units that can be sampled when a population unit originally selected can no longer be sampled. A couple of reasons why an originally selected site can no longer be sampled include its location being physically inaccessible or it is on private land that the researcher does not have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping between two-dimensional and one-dimensional space that preserves proximity relationships. First, the bounding box of the domain is split up into four distinct, equally sized squares called level-one cells. Each level-one cell is randomly assigned a level-one address of 0, 1, 2, or 3. The set of level-one cells is denoted by $\mathcal{A}_1$ and defined as $\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}$. Within each level-one cell, the inclusion probability for each population unit (which is pre-specified) is summed, and if any of these sums exceed one, a second level of cells is added. Then each level-one cell is split into four distinct, equally sized squares called level-two cells. Each level-two cell is randomly assigned a level-two address of 0, 1, 2, or 3. The set of level-two cells is denoted by $\mathcal{A}_2$ and defined as $\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}$. The inclusion probabilities within each level-two cell are summed, and if any of these sums exceed one, a third level of cells is added. This process continues for $k$ steps, until all level-$k$ cells have inclusion probability sums no larger than one. Then $\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}$. Figure$~$\ref{fig:grts} provides some intuition regarding the assignment of level-one and level-two cells.

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/grts_level1.jpeg}
  \caption{Assignment of level-one cells to the spatial domain. Grey circles indicate population units.}
  \label{fig:grts_level1}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/grts_level2.jpeg}
  \caption{Assignment of level-two cells to the spatial domain. Grey circles indicate population units.}
  \label{fig:grts_level2}
\end{subfigure}
\caption{Assignment of level-one and level-two cells to the spatial domain. In (a), each level-one cells is randomly given a level-one address of 0, 1, 2, or 3. In (b), each level-two cell within each level-one cell is randomly given a level-two address of 0, 1, 2, or 3.}
\label{fig:grts}
\end{figure}

After determining $\mathcal{A}_k$, the set is placed into hierarchical order. Hierarchical order is a numeric order that first sorts $\mathcal{A}_k$ by the level-one addresses from smallest to largest, then sorts $\mathcal{A}_k$ by the level-two addresses from smallest to largest, and so on. For example, $\mathcal{A}_2$ in hierarchical order is the set \linebreak $\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}$. Because hierarchical ordering sorts by level-one cells, then level-two cells, and so on, population units that have similar hierarchical addresses tend to be nearby one another in space. Next, each population unit is mapped to a one-dimensional line in hierarchical order where each population unit's inclusion probability equals its line-length. If a level-$k$ cell has multiple population units in it, they are randomly placed within the cell's respective line segment. A uniform random variable is then simulated in $[0, 1]$ and a systematic sample is selected on the line, yielding $n$ sample points for a sample size $n$. Each of these sample points falls on some population unit's line segment, and thus that population unit is selected in the sample. For further details regarding the GRTS algorithm, see @stevens2004spatially.

After selecting a sample and collecting data, unbiased estimates of population means and totals can be obtained using the Horvitz-Thompson estimator [@horvitz1952generalization]. If $\tau$ is a population total, the Horvitz-Thompson estimator for $\tau$, denoted by $\hat{\tau}_{ht}$, is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align}
where $z_i$ is the value of the $i$th population unit in the sample, $\pi_i$ is the inclusion probability of the $i$th population unit in the sample, and $n$ is the sample size. An estimate of the population mean is obtained by dividing $\hat{\tau}_{ht}$ by $N$, the number of population units.

@@

Finite population block kriging (FPBK) is a model-based approach that expands the geostatistical Kriging framework to the finite population setting [@verhoef2008spatial]. Instead of developing inference based on a specific sampling design, we assume the data are generated by a spatial stochastic process. We summarize some of the basic principles of FPBK next -- see @verhoef2008spatial for technical details and see @higham2021adjusting for an extension to cases of imperfect detection among population units. Let ${\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}$ be an $N \times 1$ response vector at locations $s_1$, $s_2$, . . . , $s_N$ that can be measured at the $N$ population units. Suppose we want to use a sample to predict some linear function of the response variable, $f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}$, where $\mathbf{b}^\prime$ is a $1 \times N$ vector of weights (e.g, the population mean is represented by a weights vector whose elements all equal $1 / N$). Denoting quantities that are part of the sampled population units with a subscript \emph{s} and quantities that are part of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes $\bm{\delta}$ in Equation$~$\eqref{equation:Zmarginal} has mean-zero and a spatial dependence structure that can be modeled using a covariance function. This covariance function is commonly assumed to be non-negative, second-order stationary (depending only on the separation vector (e.g., distance) between population units), isotropic (independent of direction), and decays with distance between population units [@cressie1993statistics]. Henceforth, it is implied that we have made these same assumptions regarding $\bm{\delta}$. @chiles1999geostatistics, pp. 80-93 discuss covariance functions that are not second-order stationary, not isotropic, or not either. A variety of flexible covariance functions can be used to model $\bm{\delta}$ [@cressie1993statistics]; one example is the exponential covariance function (@cressie1993statistics provides a thorough list of spatial covariance functions). The $i,j$th element of the exponential covariance matrix, $\cov(\bm{\delta})$, is
\mbox{}
\begin{align}\label{equation:expcov}
\cov(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align}
where $\sigma^2_{1}$ is the variance parameter that quantifies the spatially dependent (correlated) variability, $\sigma^2_{2}$ is the variance parameter the quantifies that spatially independent (not correlated) variability, $\phi$ is the distance parameter that measures the distance-decay rate of the covariance, and $h_{i,j}$ is the Euclidean distance between population units $i$ and $j$. In geostatistical literature, $\sigma^2_{1}$ is called the partial sill, $\sigma^2_{2}$ is called the nugget, and $\phi$ is called the range. We denote $\bm{\theta}$ as the vector of covariance parameters that composes $\bm{\delta}$. In Equation$~$\ref{equation:expcov}, $\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}$.

The parameters in Equation$~$\ref{equation:Zmarginal} can be estimated using a variety of techniques, but we focus on using restricted maximum likelihood (REML) [@patterson1971recovery; @harville1977maximum; @wolfinger1994computing]. REML is preferred over maximum likelihood (ML) because ML estimates can be badly biased for small sample sizes, due to the fact that ML makes no adjustment for the simultaneous estimation of $\bm{\beta}$ and $\bm{\delta}$ [@patterson1971recovery].  Minus twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation}
where $\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s$ and $\bm{\Sigma}_{ss}$  is the covariance matrix of the sampled sites. Minimizing Equation$~$\ref{equation:reml} yields $\bm{\hat{\theta}}_{reml}$, the REML estimates of $\bm{\delta}$. Then $\bm{\hat{\beta}}_{reml}$, the REML estimate of $\bm{\beta}$, is given by $(\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s$, where $\bm{\hat{\Sigma}}_{ss}$ is $\bm{\Sigma}_{ss}$ evaluated at $\bm{\hat{\theta}}_{reml}$.

@@

We used simulated data to compare the sampling-inference combinations across many realized populations from the same data-generating stochastic process. With the simulated data, we were in control of the data-generating stochastic process and the random sampling process. We then used real data from the 2012 National Lakes Assessment to compare the sampling-inference combinations within a single realized population (which is typically the case in reality). With the real data, we were in control of only the random sampling process.

@@

The United States Environmental Protection Agency (USEPA), states, and tribes periodically conduct National Aquatic Research Surveys (NARS) to assess the water quality of various bodies of water in the contiguous United States. One component of NARS is the National Lakes Assessment (NLA), which measures various aspects of lake health and water quality. We focus on analyzing zooplankton multi-metric indices (ZMMI) and mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA. For ZMMI, data were collected at 1035 unique lakes. At less than 10% of lakes, two ZMMI replicates were collected. These were averaged for the purposes of our study so that each lake had one measurement for ZMMI. For Hg ppb, data were collected at 995 unique lakes (there were no replicates). The ZMMI and Hg ppb data are shown as spatial maps and as histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest near the coasts, lowest in the Central United States, are relatively symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest in the Northeastern United States, lowest elsewhere, are skewed, and have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram quantifies the halved average squared differences (semivariance) of responses whose separation (distance) falls within some distance class. The spatial semivariance is closely related to the spatial covariance, and spatial semivariograms are often used to gauge the strength of spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance increases steadily with distance (meaning that observations nearby one another tend to be more similar than observations far apart from one another).

@@

Mean bias is nearly zero for all four sampling-inference combinations in all 36 scenarios, so we omit a more detailed summary of those results here. Tables for mean bias in all 36 simulation scenarios are provided in the supporting information.

We define the relative RMSE as a ratio with numerator given by the RMSE for a sampling-inference combination and the denominator given by the RMSE for SRS-DB. Relative RMSEs for the random location layout are provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 0%"), the four sampling-inference combinations have approximately equal RMSE. In these scenarios, using GRTS sampling or model-based inference does not generally increase efficiency compared to SRS-DB. When there is spatial covariance (Fig. \ref{fig:rmspe_eff}, "DRE%: 50%" and "DRE%: 90%"), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB, SRS-MB, and finally SRS-DB.  As the strength of spatial covariance increases, the gap in RMSE between SRS-DB and the other sampling-inference combinations widens. Finally we note that when there is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB, suggesting that the lack of efficiency from SRS is largely mitigated by model-based inference. These RMSE conclusions are similar to those observed in the grid location layout, so we omit a figure and discussion regarding the grid location layout here. Tables for RMSE in all 36 simulation scenarios are provided in the supporting information.
\begin{figure}
  \centering
  \includegraphics[width = 1\linewidth]{figures/rmspe_eff.jpeg}
  \caption{Relative RMSE in the simulation study for the four sampling-inference combinations and three sample sizes in the random location layout. The rows indicate the proportion of dependent error and the columns indicate the response type. The solid, black lines separate the sample sizes.}
  \label{fig:rmspe_eff}
\end{figure}
